Modelo de Simulaci\'{o}n de circulaci\'{o}n Veh\'{i}culos\\ en las calles de San Jos\'{e}, Costa Rica

### 2013-05-06 trunk/Paper/main.tex
Nuevo paper, está cachete!

%% bare_jrnl_compsoc.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE Computer
%% Society journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

@@

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
% The Computer Society usually requires 12pt for submissions.
%
\documentclass[12pt,journal,compsoc]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[12pt,journal,compsoc]{../sty/IEEEtran}

@@

% *** CITATION PACKAGES ***
%
\ifCLASSOPTIONcompsoc
  % IEEE Computer Society needs nocompress option
  % requires cite.sty v4.0 or later (November 2003)
  % \usepackage[nocompress]{cite}
\else
  % normal IEEE
  % \usepackage{cite}
\fi
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.
%
% Note that some packages require special options to format as the Computer
% Society requires. In particular, Computer Society  papers do not use
% compressed citation ranges as is done in typical IEEE papers
% (e.g., [1]-[4]). Instead, they list every citation separately in order
% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
% package with the nocompress option which is supported by cite.sty v4.0
% and later. Note also the use of a CLASSOPTION conditional provided by
% IEEEtran.cls V1.7 and later.

@@

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%\usepackage[tight,normalsize,sf,SF]{subfigure}
%\else
%\usepackage[tight,footnotesize]{subfigure}
%\fi
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. Computer Society papers
% use a larger font and \sffamily font for their captions, hence the
% additional options needed under compsoc mode. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false]{caption}
%  \usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false]{caption}
%  \usepackage[font=footnotesize]{subfig}
%\fi
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

@@

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
% This package may be useful when used in conjunction with IEEEtran.cls'
% captionsoff option. Some IEEE journals/societies require that submissions
% have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.3.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% For subfigure.sty:
% \let\MYorigsubfigure\subfigure
% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat/subfig command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

@@

\usepackage{graphicx}
\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Modelo de Simulac\'{i}on de circulaci\'{o}n Veh\'{i}culos\\ en las calles de San Jos\'{e}, Costa Rica}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc 
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

\author{Bchr. Mag Yari~Guevara Rivera,
        Msc. Eddy Miguel~Ram\'irez-Jim\'enez}% <-this % stops a space
%\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem M. Shell is with the Department
%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%GA, 30332.\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
%E-mail: see http://www.michaelshell.org/contact.html
%\IEEEcompsocthanksitem J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Modelo de simulaci\'on para el evaluar el comportamiento del tr\'ansito de San Jos\'e, Costa Rica}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Computer Society Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% The publisher's ID mark at the bottom of the page is less important with
% Computer Society journal papers as those publications place the marks
% outside of the main text columns and, therefore, unlike regular IEEE
% journals, the available text space is not reduced by their presence.
% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% or like this to get the Computer Society new two part style.
%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2007 IEEE}%
%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark (Computer Society jorunal
% papers don't need this extra clearance.)

@@

\setlength\parindent{24pt}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
One of the most common problems in Costa Rica's capital, San Jose, is the heavy traffic jams we are usually exposed to. Any possible solutions is often applied and after a rate os success or failure, it is kept or discharched. It is mandatory to have a system available in the Ministery encharged of this issue a simulation model that envolves all the characteristics of the local situations. This paper shows how this simulation system was developed according the particular needs that exist in Costa Rica and the initial results of it.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Computer Society, Probability distribution, Statistics, Simulation, Traffic Lights, Model.
\end{IEEEkeywords}}

@@

% To allow for easy dual compilation without having to reenter the
% abstract/keywords data, the \IEEEcompsoctitleabstractindextext text will
% not be used in maketitle, but will appear (i.e., to be "transported")
% here as \IEEEdisplaynotcompsoctitleabstractindextext when compsoc mode
% is not selected <OR> if conference mode is selected - because compsoc
% conference papers position the abstract like regular (non-compsoc)
% papers do!
\IEEEdisplaynotcompsoctitleabstractindextext
% \IEEEdisplaynotcompsoctitleabstractindextext has no effect when using
% compsoc under a non-conference mode.

@@

\section{Introducci\'{o}n}

@@

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{E}{n} Costa Rica a lo largo de los a\~{n}os se ha podido observar como la gran
	cantidad de veh\'{i}culos automotores saturan sus principales ciudades, en especial
	el caso de su capital, donde las calles fueron creadas a inicios del S. XX y han sufrido pocas modificaciones, salvo por la ampliación de la Avenida Juan Rafael Mora Porras. Para el 2012 y seg\'{u}n el peri\'{o}dico La Naci\'{o}n
	alrededor de 19.000 buses y 260.000 autos luchan por un espacio para poder avanzar por las calles de San Jos\'{e}. \cite{Villegas2012}\\

@@

		La gran cantidad de veh\'{i}culos automotores que circulan en las diferentes \'{a}reas
	de San Jos\'{e} provocan d\'{i}a a d\'{i}a embotellamientos que aumentan el tiempo requerido para ir de un punto a otro, as\'{i} como el consumo de combustible y por ende las emisiones de gases contaminantes.
	La situaci\'{o}n se presenta m\'{a}s alarmante cuando el mismo alcalde de San Jos\'{e}, el
	se\~{n}or Johnny Araya afirma que la cantidad de autos mencionados
	anteriormente ocupan un 70\% del espacio vial en la capital, pero \'{u}nicamente trasladan el 30\% del mill\'{o}n de personas que ingresan a San Jos\'{e} todos los d\'{i}as.\cite{Villegas2012}, lo que quiere decir que la gran mayoría de personas utilizan el transporte p\'ublico para hacer su ingreso a San Jos\'e.

@@

Existen diferentes escenarios por los cuales se puedan dar los problemas planteados anteriormente, para este trabajo se utilizar\'{a} como zona de estudio la representada en la figura \ref{fig:traficoSJ}, sin embargo, el programa será lo suficientemente flexible para poder modelar diferentes zonas de tr\'ansito.

@@

	Es aqu\'{i} donde la primera parte de este trabajo (la simulaci\'{o}n) entra en juego. Para este caso la simulaci\'{o}n s\'{o}lo comprender\'{a} escenarios normales dentro de la zona, si ningun evento o factor que altere la presencia o el flujo de veh\'{i}culos dentro de la misma. Otro aspecto importante a considerar es que el periodo a evaluarse corresponde a la franja horaria de las 4pm a las 6pm, dado que seg\'{u}n estimaciones realizadas en el MOPT, presenta una tendencia de flujo m\'{a}s constante en comparaci\'{o}n con las horas de la ma\~{n}ana donde se obtienen picos pronunciados en menor tiempo, correspondientes al traslado de personas a su lugar de trabajo.

@@

\subsection{Factores a Considerarse}
Bas\'{a}ndose en eventos y condiciones que tienden a presentarse en la mayor\'{i}a de zonas del pa\'{i}s, se determinaron los factores a ser tomados en cuenta para la realizaci\'{o}n de este experimento total y a ser empleado en conjunto con las simulaciones. Cada factor consta de dos a tres niveles a ser tomados en cuenta, dichos niveles corresponden a la dificultad que el evento conlleva para los otros veh\'{i}culos dentro de la zona. Los factores se explican a continuaci\'{o}n.

@@

	Cabe mencionar que para este caso s\'{o}lo se considerar\'{a}n casos de paradas permanentes, por el hecho de que realizar evaluaciones de otros resulta bastante difícil debido al grado de variación que se presenta, tal es el caso de el parqueo temporal de vehículos el cual dependerá de cada persona, según las habilidades y experiencia con las que cuenten para tardar más o menos tiempo en estacionar un vehículo.

@@

\subsubsection{Condiciones de los sem\'aforos}
Uno de los principales factores que deben de considerarse es el estado de los semáforos, indicando qu\'e calles y avenidas est\'an avanzando y cu\'ales est\'an detenidas. Este factor debe de poder ser controlado de manera din\'amica, para poder buscar mejoras posteriores, al ser el \'unico sobre el cual se tiene control externo, a diferencia de los mencionados anteriormente.

@@

\subsection{Entrada de veh\'iculos}
La hora (con segundos) de llegada de veh\'iculos se obtiene a partir de una distribuci\'on de Poisson, cuya media var\'ia de acuerdo con las horas del d\'ia. Donde la media de menor valor corresponde con el per\'iodo comprendido entre las cuatro de la tarde y las siente de la noche, con una media de 5. Esto quiere decir que en promedio, entra un autom\'ovil cada cinco segundos por cada una de las posibles entradas. Es interesante como una muestra similar se puede apreciar entre las siete y ocho de la ma\~nana, sin embargo, en este evento, los autos transitan con mayor premura, debido a que muchos van a hacer su ingreso a su lugar de trabajo lo que agiliza significativamente el tr\'ansito.

@@

	Por otro lado, la demora total corresponde a los tiempos acumulados para los carros, que se generen por eventos que causen atrasos a estos como el caso de obstrucciones en las vía o que las calles han llegado al máximo de su capacidad y por lo tanto tienen que esperar fuera de estas para poder ingresar y continuar con la ruta. Finalmente se cuenta con la posici\'on de los automotores en cada calle, para este caso debe tenerse en cuenta que la posici\'on es \'unica para cada calle y al momento de que se pase a una nueva la misma ser\'a reiniciada al valor m\'aximo posible dependiendo de la capacidad de la calle.

@@

\section{Trabajo Futuro}
Posterior a la realizaci\'{o}n del modelo de simulaci\'{o}n, es importante encontrar un mecanismo que agilice el control del tr\'ansito. En el proceso pretende lograr un modelo de red neuronal artificial que pueda ser implementada dentro de este red de sem\'{a}foros y el cual permita que cada sem\'{a}foro se comunique con otros (hermanos) localizados dentro de un rango de \'{a}rea considerable de forma que cada uno pueda tener una idea lo que ocurre en su entorno y no \'{u}nicamente la informaci\'{o}n que puede ser recolectada por medio de los sensores en la intersecci\'{o}n administrada por cada sem\'{a}foro.

@@

>>>>

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

@@

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

@@

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

@@

@@

\usepackage{graphicx}
\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Modelo de Simulaci\'{o}n de circulaci\'{o}n Veh\'{i}culos\\ en las calles de San Jos\'{e}, Costa Rica}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Mag Yari Guevara-Rivera, Bchr.}
\IEEEauthorblockA{Escuela de Ingenier\'ia en Sistemas\\
Universidad Nacional de Costa Rica\\
Heredia, Costa Rica\\
magyarigr@gmail.com}
\and
\IEEEauthorblockN{Eddy Ram\'irez-Jim\'enez, Msc.}
\IEEEauthorblockA{Escuela de Ingenier\'ia en Sistemas\\
Universidad Nacional de Costa Rica\\
Heredia, Costa Rica\\
eddy.ramirez.jimenez@gmail.com}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

@@

@@

\begin{abstract}
One of the most common problems in Costa Rica's capital, San Jose, is the heavy traffic jams we are usually exposed to. Any possible solutions is often applied and after a rate os success or failure, it is kept or discharched. It is mandatory to have a system available in the Ministery encharged of this issue a simulation model that envolves all the characteristics of the local situations. This paper shows how this simulation system was developed according the particular needs that exist in Costa Rica and the initial results of it.\\ \\Keywords: Computer Society, Probability distribution, Statistics, Simulation, Traffic Lights, Model.
\end{abstract}
%\begin{IEEEkeywords}
%Computer Society, Probability distribution, Statistics, %Simulation, Traffic Lights, Model.
%\end{IEEEkeywords}}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

% no keywords

@@

\section{Introducci\'on}
Un modelo de simulaci\'on es un sistema de computaci\'on que se encarga de representar por software la din\'amica de un proceso complejo, donde normalmente otro tipo de an\'alisis resulta muy costoso o en algunas ocasiones resulta imposible obtener resultados fidedignos en un tiempo razonable [1].

El el caso de Costa Rica, se presentan ciertas caracter\'isticas dadas por la sociedad y la forma en como las personas manejan en los estrechos y poco reglamentados espacios de nuestra capital, donde las llamadas \textit{presas} o embotellamientos son de todos los d\'ias.

Este proyecto desarrolla un sistema de simulaci\'on basado en las caracter\'isticas propias de nuestra regi\'on particular y las fuentes para hacerlo fueron los entes gubernamentales encargados del tr\'ansito en Costa Rica.

Este art\'iculo cuenta con 8 secciones divididas de la sigueinte manera:
\begin{enumerate}
\item La segunda trata sobre las caracter\'isticas particulares que tiene la capital de Costa Rica con respecto a la circulaci\'on de veh\'iculos. 
\item La tercera parte sobre el problema y el detalle de la zona que se ve afectada mayormente. 
\item La cuarta son las consideraciones que se tomaron en cuenta en la implementaci\'on del modelo, es decir el universo de discurso. 
\item La quinta secci\'on hace referencia a las distribuciones de probabilidad que se utilizaron para generar valores aleatorios que se asemejen con la situaci\'on en Costa Rica.
\item La sexta sección trata sobre los resultados y  valores obtenidos del simulador
\item La s\'eptima es el trabajo futuro sobre proyectos que utilizan como base al simulador
\item La octava y \'ultima muestra las conclusiones

\end{enumerate}

@@

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{E}{n} Costa Rica a lo largo de los a\~{n}os se ha podido observar como la gran
	cantidad de veh\'{i}culos automotores saturan sus principales ciudades, en especial
	el caso de su capital, donde las calles fueron creadas a inicios del S. XX y han sufrido pocas modificaciones, salvo por la ampliaci\'on de la Avenida Juan Rafael Mora Porras. Para el 2012 y seg\'{u}n el peri\'{o}dico La Naci\'{o}n
	alrededor de 19.000 buses y 260.000 autos luchan por un espacio para poder avanzar por las calles de San Jos\'{e} \cite{Villegas2012}\\ .

@@

		La gran cantidad de veh\'{i}culos automotores que circulan en las diferentes \'{a}reas
	de San Jos\'{e} provocan d\'{i}a a d\'{i}a embotellamientos que aumentan el tiempo requerido para ir de un punto a otro, as\'{i} como el consumo de combustible y por ende las emisiones de gases contaminantes.
	La situaci\'{o}n se presenta m\'{a}s alarmante cuando el mismo alcalde de San Jos\'{e}, el
	se\~{n}or Johnny Araya afirma que la cantidad de autos mencionados
	anteriormente ocupan un 70\% del espacio vial en la capital, pero \'{u}nicamente trasladan el 30\% del mill\'{o}n de personas que ingresan a San Jos\'{e} todos los d\'{i}as.\cite{Villegas2012}, lo que quiere decir que la gran mayor\'ia de personas utilizan el transporte p\'ublico para hacer su ingreso a San Jos\'e.

@@

Existen diferentes escenarios por los cuales se puedan dar los problemas planteados anteriormente, para este trabajo se utilizar\'{a} como zona de estudio la representada en la figura \ref{fig:traficoSJ}, sin embargo, el programa ser\'a lo suficientemente flexible para poder modelar diferentes zonas de tr\'ansito.

@@

	Es aqu\'{i} donde la primera parte de este trabajo (la simulaci\'{o}n) se pone en funcionamiento. Para este caso la simulaci\'{o}n s\'{o}lo comprender\'{a} escenarios normales dentro de la zona, si ningun evento o factor que altere la presencia o el flujo de veh\'{i}culos dentro de la misma. Otro aspecto importante a considerar es que el periodo a evaluarse corresponde a la franja horaria de las 4pm a las 6pm, dado que seg\'{u}n estimaciones realizadas en el MOPT, presenta una tendencia de flujo m\'{a}s constante en comparaci\'{o}n con las horas de la ma\~{n}ana donde se obtienen picos pronunciados en menor tiempo, correspondientes al traslado de personas a su lugar de trabajo.

@@

\subsection{Factores a Considerarse}
Los factores se obtienen con base en los eventos y condiciones que tienden a presentarse en la mayor\'{i}a de zonas del pa\'{i}s.

Cada factor consta de dos a tres niveles a ser tomados en cuenta, dichos niveles implica un diferente nivel de dificultad para el avance normal de los otros veh\'{i}culos dentro de la zona. Los factores se explican a continuaci\'{o}n.

@@

	Cabe mencionar que para este caso s\'{o}lo se considerar\'{a}n casos de paradas permanentes, por el hecho de que realizar evaluaciones de otros resulta bastante dif\'icil debido al grado de variaci\'on que se presenta, tal es el caso de el parqueo temporal de veh\'iculos el cual depender\'a de cada persona, seg\'un las habilidades y experiencia con las que cuenten para tardar m\'as o menos tiempo en estacionar un veh\'iculo.

@@

\subsubsection{Condiciones de los sem\'aforos}
Uno de los principales factores que deben de considerarse es el estado de los sem\'aforos, indicando qu\'e calles y avenidas est\'an avanzando y cu\'ales est\'an detenidas. Este factor debe de poder ser controlado de manera din\'amica, para poder buscar mejoras posteriores, al ser el \'unico sobre el cual se tiene control externo, a diferencia de los mencionados anteriormente.

@@

\subsection{Entrada de veh\'iculos}
La hora (con segundos) de llegada de veh\'iculos se obtiene a partir de una distribuci\'on de Poisson, cuya media var\'ia de acuerdo con las horas del d\'ia. Donde la media de menor valor corresponde con el per\'iodo comprendido entre las cuatro de la tarde y las siete de la noche, con una media de 5. Esto quiere decir que en promedio, entra un autom\'ovil cada cinco segundos por cada una de las posibles entradas. Es interesante como una muestra similar se puede apreciar entre las siete y ocho de la ma\~nana, sin embargo, en este evento, los autos transitan con mayor premura, debido a que muchos van a hacer su ingreso a su lugar de trabajo lo que agiliza significativamente el tr\'ansito.

@@

	Por otro lado, la demora total corresponde a los tiempos acumulados para los carros, que se generen por eventos que causen atrasos a estos como el caso de obstrucciones en las v\'ia o que las calles han llegado al m\'aximo de su capacidad y por lo tanto tienen que esperar fuera de estas para poder ingresar y continuar con la ruta. Finalmente se cuenta con la posici\'on de los automotores en cada calle, para este caso debe tenerse en cuenta que la posici\'on es \'unica para cada calle y al momento de que se pase a una nueva la misma ser\'a reiniciada al valor m\'aximo posible dependiendo de la capacidad de la calle.

@@

\section{Trabajo Futuro}

Resulta destacable hay algunos datos que se obtienen que no pueden verificarse su exactitud debido a que a la fecha, ninguna entidad gubernamental o privada posee datos sobre ciertos tipos de comportamientos extraños en los ciudadanos, como el caso de un choque leve donde se arreglan por su parte y no se hace el reporte a la oficina de tr\'ansito. Por lo que el modelo puede ser inexacto ante eventualidades como esta, de modo que si se instalan sensores que indiquen comportamientos extraños, entonces se puede retroalimentar al simulador.

Posterior a la realizaci\'{o}n del modelo de simulaci\'{o}n, es importante encontrar un mecanismo que agilice el control del tr\'ansito. En el proceso pretende lograr un modelo de red neuronal artificial que pueda ser implementada dentro de este red de sem\'{a}foros y el cual permita que cada sem\'{a}foro se comunique con otros (hermanos) localizados dentro de un rango de \'{a}rea considerable de forma que cada uno pueda tener una idea lo que ocurre en su entorno y no \'{u}nicamente la informaci\'{o}n que puede ser recolectada por medio de los sensores en la intersecci\'{o}n administrada por cada sem\'{a}foro.

@@

Las particularidades de la sociedad costarricense aunado a un espacio vial reducido que por lo general se encuentra a\'un m\'as reducido por diversas razones, se ve reflejado de manera muy similar a la realidad en el sistema de simulaci\'on

Las variables obtenidas de los expertos, permitieron que el modelo se completara y se pudiera ajustar a los detalles que satisfacen las condiciones precisas y los resultados obtenidos se pueden ajustar para diferentes horarios del d\'ia.
